---
title: "Session 1.2: Statistical Conclusion Validity & Power"
subtitle: "HPM 883: Advanced Quantitative Methods"
author: "Sean Sylvia"
date: "January 21, 2026"
date-format: long
format:
  revealjs:
    theme: [default, ../style/gillings.scss]
    slide-number: true
    transition: fade
    progress: true
    chalkboard: true
    smaller: true
    scrollable: true
    logo: ../images/logo.png
    footer: "HPM 883 | Session 1.2"
    toc: true
    toc-depth: 1
    preview-links: auto
    incremental: false
execute:
  echo: true
  warning: false
  message: false
---

```{r setup}
#| include: false
library(tidyverse)
library(knitr)
library(kableExtra)
library(DeclareDesign)
library(estimatr)
set.seed(883)
```

# Introduction

## From Session 1.1: What We Learned

::: {.callout-note}
## Randomization Solves the Identification Problem
Under randomization: $\mathbb{E}[Y|D=1] - \mathbb{E}[Y|D=0] = \text{ATE}$

We can identify the causal effect without selection bias.
:::

. . .

**But identification isn't everything...**

::: {.fragment}
- Can we **detect** the effect if it exists?
- How **precise** will our estimate be?
- What **sample size** do we need?
:::

## Two Types of Validity

::: {.columns}
::: {.column width="50%"}
### Internal Validity
**Is the estimate unbiased?**

- Does $\mathbb{E}[\hat{\tau}] = \tau$?
- Randomization ensures this
- Session 1.1 focused here
:::

::: {.column width="50%"}
### Statistical Conclusion Validity
**Can we reliably detect effects?**

- Will we reject $H_0$ when effect exists?
- Depends on power, sample size, variance
- Today's focus
:::
:::

. . .

::: {.callout-tip}
## The Goal
An experiment with **both**: unbiased estimation (internal validity) AND adequate precision (statistical conclusion validity).
:::

## Session Roadmap

| Section | Topic | Time |
|---------|-------|------|
| 1 | Type I & Type II Errors | 10 min |
| 2 | Power Analysis | 20 min |
| 3 | Power by Simulation | 20 min |
| 4 | Variance Reduction | 15 min |
| 5 | Clustering & Multiple Testing | 10 min |

# Type I and Type II Errors

## The Hypothesis Testing Framework

::: {.callout-note}
## Setup
- **Null hypothesis:** $H_0: \tau = 0$ (no treatment effect)
- **Alternative:** $H_1: \tau \neq 0$ (effect exists)
- **Decision rule:** Reject $H_0$ if p-value $< \alpha$
:::

. . .

**Two ways to be wrong:**

| Truth | Reject $H_0$ | Fail to Reject $H_0$ |
|-------|--------------|----------------------|
| $H_0$ true (no effect) | **Type I Error** ($\alpha$) | Correct |
| $H_1$ true (effect exists) | Correct | **Type II Error** ($\beta$) |

## Visualizing the Errors

```{r error-viz}
#| fig-height: 4
#| fig-width: 10
#| echo: false

# Simulate distributions under H0 and H1
n <- 200
se <- 0.15
true_effect <- 0.5

x <- seq(-0.6, 1.2, length.out = 1000)
h0_dist <- dnorm(x, mean = 0, sd = se)
h1_dist <- dnorm(x, mean = true_effect, sd = se)

# Critical value
alpha <- 0.05
crit <- qnorm(1 - alpha/2, mean = 0, sd = se)

df <- tibble(x = x, H0 = h0_dist, H1 = h1_dist) %>%
  pivot_longer(-x, names_to = "hypothesis", values_to = "density")

ggplot() +
  # H0 distribution
  geom_area(data = filter(df, hypothesis == "H0" & x > crit),
            aes(x = x, y = density), fill = "red", alpha = 0.4) +
  geom_line(data = filter(df, hypothesis == "H0"),
            aes(x = x, y = density), color = "darkblue", linewidth = 1) +
  # H1 distribution
  geom_area(data = filter(df, hypothesis == "H1" & x < crit),
            aes(x = x, y = density), fill = "orange", alpha = 0.4) +
  geom_line(data = filter(df, hypothesis == "H1"),
            aes(x = x, y = density), color = "darkgreen", linewidth = 1) +
  # Critical value
  geom_vline(xintercept = crit, linetype = "dashed", color = "black") +
  annotate("text", x = 0, y = max(h0_dist) * 1.1, label = "H₀: τ = 0", color = "darkblue") +
  annotate("text", x = true_effect, y = max(h1_dist) * 1.1, label = "H₁: τ = 0.5", color = "darkgreen") +
  annotate("text", x = crit + 0.08, y = 0.5, label = "α (Type I)", color = "red") +
  annotate("text", x = crit - 0.15, y = 0.5, label = "β (Type II)", color = "orange") +
  labs(x = "Estimated Effect", y = "Density",
       title = "Type I and Type II Error Regions") +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

## Power: The Probability of Detection

::: {.callout-important}
## Definition
**Power** = $1 - \beta$ = P(reject $H_0$ | $H_1$ is true)

The probability of detecting an effect **when it exists**.
:::

. . .

**Conventional targets:**

| Field | Minimum Power |
|-------|---------------|
| Medical trials | 80% (often 90%) |
| Social science | 80% |
| Psychology replication | 95%+ recommended |

. . .

::: {.callout-warning}
## Underpowered Studies
- High false negative rate
- Estimates conditional on significance are biased upward
- Wastes resources, unethical if treatment could help
:::

## What Determines Power?

Power increases with:

::: {.incremental}
1. **Larger effect size** ($\tau$) — easier to detect big effects
2. **Larger sample size** ($n$) — reduces sampling variance
3. **Lower outcome variance** ($\sigma^2$) — cleaner signal
4. **Higher significance level** ($\alpha$) — accept more Type I risk
5. **One-sided test** — (but usually inappropriate)
:::

. . .

::: {.callout-note}
## The Key Tradeoff
We control $n$, $\alpha$, and design (which affects variance).

Effect size is determined by nature. Sample size is limited by budget.
:::

# Power Analysis

## The Variance of the Treatment Effect Estimator

For a simple RCT with difference-in-means:

$$\hat{\tau} = \bar{Y}_1 - \bar{Y}_0$$

. . .

**Variance under randomization:**

$$\text{Var}(\hat{\tau}) = \frac{\sigma_1^2}{n_1} + \frac{\sigma_0^2}{n_0}$$

where:

- $\sigma_1^2$ = variance of $Y(1)$
- $\sigma_0^2$ = variance of $Y(0)$
- $n_1, n_0$ = sample sizes in each arm

## Simplifying with Equal Allocation

If $n_1 = n_0 = n/2$ and $\sigma_1^2 = \sigma_0^2 = \sigma^2$:

$$\text{Var}(\hat{\tau}) = \frac{2\sigma^2}{n/2} = \frac{4\sigma^2}{n}$$

. . .

**Standard error:**

$$\text{SE}(\hat{\tau}) = \sigma\sqrt{\frac{4}{n}} = \frac{2\sigma}{\sqrt{n}}$$

. . .

::: {.callout-tip}
## Key Insight
Standard error decreases with $\sqrt{n}$, not $n$.

To halve SE, you need 4× the sample size!
:::

## Deriving the Power Formula

**Reject $H_0$ when:** $|\hat{\tau}|/\text{SE} > z_{1-\alpha/2}$

. . .

**Under $H_1$** (true effect = $\tau$):

$$\frac{\hat{\tau}}{\text{SE}} \sim N\left(\frac{\tau}{\text{SE}}, 1\right)$$

. . .

**Power:**

$$\text{Power} = \Phi\left(\frac{\tau}{\text{SE}} - z_{1-\alpha/2}\right) + \Phi\left(-\frac{\tau}{\text{SE}} - z_{1-\alpha/2}\right)$$

For practical purposes (one-sided approximation):

$$\text{Power} \approx \Phi\left(\frac{\tau}{\text{SE}} - z_{1-\alpha/2}\right)$$

## Minimum Detectable Effect (MDE)

::: {.callout-important}
## Definition
**MDE** = smallest effect detectable with power $1-\beta$ at significance $\alpha$
:::

. . .

**Formula:**

$$\text{MDE} = (z_{1-\alpha/2} + z_{1-\beta}) \times \text{SE}$$

For a two-arm RCT with equal allocation:

$$\text{MDE} = (z_{1-\alpha/2} + z_{1-\beta}) \times \sigma\sqrt{\frac{4}{n}}$$

. . .

**With $\alpha = 0.05$, power = 80%:**

$$\text{MDE} \approx 2.8 \times \sigma\sqrt{\frac{4}{n}} = \frac{5.6\sigma}{\sqrt{n}}$$

## MDE Calculator

```{r mde-calc}
#| echo: true
calculate_mde <- function(n, sigma, alpha = 0.05, power = 0.80) {
  z_alpha <- qnorm(1 - alpha/2)
  z_beta <- qnorm(power)
  se <- sigma * sqrt(4/n)
  mde <- (z_alpha + z_beta) * se
  return(mde)
}

# Example: HbA1c study
sigma <- 1.2  # SD of HbA1c
n_total <- 400

mde <- calculate_mde(n = n_total, sigma = sigma)
cat("With N =", n_total, "and SD =", sigma, "\n")
cat("MDE =", round(mde, 3), "% HbA1c\n")
```

## Power Curve: Sample Size vs Power

```{r power-curve}
#| fig-height: 4
#| fig-width: 9
#| echo: false

# Calculate power for different sample sizes
power_data <- tibble(
  n = seq(50, 600, by = 25)
) %>%
  mutate(
    se = 1.2 * sqrt(4/n),
    power = pnorm(0.5/se - qnorm(0.975))  # effect = 0.5
  )

ggplot(power_data, aes(x = n, y = power)) +
  geom_line(linewidth = 1.2, color = "#4a90a4") +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "red") +
  geom_point(data = filter(power_data, abs(power - 0.8) == min(abs(power - 0.8))),
             size = 4, color = "red") +
  annotate("text", x = 500, y = 0.82, label = "80% Power", color = "red") +
  labs(
    x = "Total Sample Size",
    y = "Statistical Power",
    title = "Power by Sample Size",
    subtitle = "Effect = 0.5, SD = 1.2, α = 0.05"
  ) +
  theme_minimal(base_size = 14) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent)
```

# Power by Simulation

## Why Simulate?

Analytical formulas work for simple designs, but...

::: {.incremental}
- **Blocking** changes the variance formula
- **Clustering** introduces ICC
- **Covariate adjustment** reduces variance (but by how much?)
- **Complex estimands** (LATE, ATT) have different variances
- **Nonlinear models** (logistic) don't have clean formulas
:::

. . .

::: {.callout-tip}
## The Solution
**Simulate your experiment many times** and see how often you reject $H_0$.

Power = proportion of simulations where you reject.
:::

## Introduction to DeclareDesign

::: {.callout-note}
## The MIDA Framework
**M**odel → **I**nquiry → **D**ata Strategy → **A**nswer Strategy

A complete research design declaration.
:::

```{r dd-intro}
#| echo: true
#| eval: false
library(DeclareDesign)

# 1. Model: How does the world work?
model <- declare_model(...)

# 2. Inquiry: What do we want to know?
inquiry <- declare_inquiry(...)

# 3. Data: How do we collect data?
data_strategy <- declare_assignment(...) + declare_measurement(...)

# 4. Answer: How do we estimate?
answer <- declare_estimator(...)

# Combine into a design
design <- model + inquiry + data_strategy + answer
```

## Building a Design Step-by-Step

```{r dd-design}
#| echo: true
# Parameters
N <- 400
effect <- -0.5
sigma <- 1.2

# Model: Potential outcomes
model <- declare_model(
  N = N,
  U = rnorm(N, 0, sigma),
  Y_Z_0 = 8.5 + U,
  Y_Z_1 = 8.5 + effect + U
)

# Inquiry: ATE
inquiry <- declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0))

# Data: Random assignment and measurement
assignment <- declare_assignment(Z = complete_ra(N, prob = 0.5))
measurement <- declare_measurement(Y = reveal_outcomes(Y ~ Z))

# Answer: Difference-in-means
estimator <- declare_estimator(Y ~ Z, inquiry = "ATE",
                                .method = lm_robust, label = "DIM")

# Combine
design <- model + inquiry + assignment + measurement + estimator
```

## Diagnosing the Design

```{r dd-diagnose}
#| echo: true
#| cache: true
# Run 500 simulations
diagnosis <- diagnose_design(design, sims = 500)

# Key diagnosands
diagnosis$diagnosands_df %>%
  select(estimator, bias, sd_estimate, power, coverage) %>%
  kable(digits = 3)
```

. . .

::: {.callout-note}
## Interpretation
- **Bias ≈ 0**: Difference-in-means is unbiased ✓
- **Power**: Probability of rejecting $H_0$ when effect exists
- **Coverage ≈ 0.95**: CI contains true value 95% of the time ✓
:::

## Comparing Designs: Blocked vs Simple

```{r dd-blocked}
#| echo: true
#| cache: true
# Add clinic blocking variable
model_blocked <- declare_model(
  N = N,
  clinic = sample(1:20, N, replace = TRUE),
  clinic_effect = rnorm(20, 0, 0.4)[clinic],
  U = rnorm(N, 0, 1.0),
  Y_Z_0 = 8.5 + clinic_effect + U,
  Y_Z_1 = 8.5 + effect + clinic_effect + U
)

# Blocked assignment
assign_blocked <- declare_assignment(Z = block_ra(blocks = clinic))

# Estimator with clinic fixed effects
est_blocked <- declare_estimator(Y ~ Z + factor(clinic), inquiry = "ATE",
                                  term = "Z", .method = lm_robust,
                                  label = "Blocked + FE")

design_blocked <- model_blocked + inquiry + assign_blocked + measurement + est_blocked
design_simple <- model_blocked + inquiry + assignment + measurement + estimator

comparison <- diagnose_designs(
  list(Simple = design_simple, Blocked = design_blocked), sims = 500
)
```

## Comparison Results

```{r dd-compare-results}
#| echo: false
comparison$diagnosands_df %>%
  select(design, estimator, sd_estimate, power, coverage) %>%
  kable(digits = 3, caption = "Simple vs Blocked Randomization")
```

. . .

::: {.callout-tip}
## Why Blocking Helps
Blocking removes between-clinic variance from the error term.

Lower variance → smaller SE → more power!
:::

# Variance Reduction

## Lin (2013): Proper Covariate Adjustment

::: {.callout-important}
## The Problem
Freedman (2008) showed that ANCOVA can be biased in finite samples.

Lin (2013) provided the fix.
:::

. . .

**Lin's Recommendation:**

1. **Center** covariates at their sample means
2. **Include** covariate × treatment interactions
3. **Use HC2** robust standard errors

$$Y_i = \alpha + \tau Z_i + \beta (X_i - \bar{X}) + \gamma Z_i(X_i - \bar{X}) + \varepsilon_i$$

## Why Lin Adjustment Works

```{r lin-demo}
#| echo: true
# Design with baseline covariate
model_cov <- declare_model(
  N = 400,
  X = rnorm(N, 8.5, 0.5),  # Baseline HbA1c
  U = 0.7 * (X - 8.5) + rnorm(N, 0, 0.8),  # Correlated error
  Y_Z_0 = 8.5 + U,
  Y_Z_1 = 8.5 + effect + U
)

# Unadjusted
est_unadj <- declare_estimator(Y ~ Z, inquiry = "ATE",
                                .method = lm_robust, label = "Unadjusted")

# Lin-adjusted
est_lin <- declare_estimator(Y ~ Z * I(X - mean(X)), inquiry = "ATE",
                              term = "Z", .method = lm_robust, label = "Lin")

design_unadj <- model_cov + inquiry + assignment + measurement + est_unadj
design_lin <- model_cov + inquiry + assignment + measurement + est_lin
```

## Lin Adjustment Results

```{r lin-results}
#| echo: false
#| cache: true
lin_comparison <- diagnose_designs(
  list(Unadjusted = design_unadj, Lin = design_lin), sims = 500
)

lin_comparison$diagnosands_df %>%
  select(design, bias, sd_estimate, power, coverage) %>%
  kable(digits = 3, caption = "Effect of Lin Adjustment")
```

. . .

::: {.callout-note}
## Key Takeaway
Lin adjustment **never hurts** (asymptotically) and often substantially increases power.

**Always pre-register your adjustment strategy!**
:::

# Clustering and Multiple Testing (Preview)

## Cluster Randomization: The Problem

::: {.callout-warning}
## When Units Aren't Independent
- Randomize clinics, not patients
- Students within classrooms
- Villages in development RCTs
:::

. . .

**Effective sample size:**

$$n_{\text{eff}} = \frac{n}{1 + (m-1)\rho}$$

where:

- $n$ = total observations
- $m$ = cluster size
- $\rho$ = intra-cluster correlation (ICC)

## Design Effect

::: {.callout-important}
## Design Effect (DEFF)
$$\text{DEFF} = 1 + (m - 1) \times \rho$$

Multiplier on required sample size due to clustering.
:::

. . .

**Example:**

- 20 clinics, 20 patients each ($n = 400$, $m = 20$)
- ICC = 0.05

$$\text{DEFF} = 1 + (20-1)(0.05) = 1.95$$

Effective $n \approx 205$, not 400!

. . .

::: {.callout-tip}
**We'll cover clustering in depth in Session 1.3.**
:::

## Multiple Testing: A Brief Preview

::: {.columns}
::: {.column width="50%"}
**Family-wise Error Rate (FWER)**

P(at least one false positive)

Bonferroni: $\alpha_{\text{adj}} = \alpha / k$
:::

::: {.column width="50%"}
**False Discovery Rate (FDR)**

Expected proportion of false positives

Benjamini-Hochberg: control FDR at $q$
:::
:::

. . .

::: {.callout-note}
## Best Practice
**Pre-specify a single primary outcome** in your analysis plan.

Secondary outcomes should be explicitly labeled as exploratory.
:::

# Wrap-Up

## Key Takeaways

::: {.incremental}
1. **Statistical conclusion validity** asks: can we detect effects with adequate precision?
2. **Power** = P(reject $H_0$ | effect exists) — aim for ≥ 80%
3. **MDE formula**: $(z_\alpha + z_\beta) \times \text{SE}$
4. **Power by simulation** handles complex designs DeclareDesign can't solve analytically
5. **Lin adjustment** reduces variance without introducing bias
6. **Clustering** reduces effective sample size — account for it!
:::

## For the Remote Lab (Monday)

::: {.callout-note}
## Lab 1: Power Analysis with DeclareDesign
You will:

1. Declare a complete experimental design (MIDA)
2. Diagnose power across sample sizes
3. Compare simple vs blocked randomization
4. Calculate MDEs
5. Design a smoking cessation trial

**Submit by Feb 2** via Gradescope.
:::

## Questions?

::: {.callout-note appearance="minimal"}
**Office Hours:** Tuesdays 2-3pm or by appointment

**Slack:** #hpm883-questions

**Readings for Session 1.3:** Duflo et al. (2007), Bruhn & McKenzie (2009)
:::

# Appendix {visibility="uncounted"}

## References {visibility="uncounted"}

- Chernozhukov et al. (2025). Chapter 2: Causal Inference via Randomized Experiments.
- Lin, W. (2013). Agnostic notes on regression adjustments. *Statistics, Politics, and Policy*.
- Blair et al. (2019). Declaring and Diagnosing Research Designs. *APSR*.
- Freedman, D. (2008). On regression adjustments in experiments. *Annals of Applied Statistics*.

## Power Formula Derivation {visibility="uncounted"}

Under $H_0: \tau = 0$:
$$T = \frac{\hat{\tau}}{\text{SE}} \sim N(0, 1)$$

Under $H_1: \tau = \tau^*$:
$$T = \frac{\hat{\tau}}{\text{SE}} \sim N\left(\frac{\tau^*}{\text{SE}}, 1\right)$$

Power = P(|T| > $z_{1-\alpha/2}$ | $H_1$)

$$= \Phi\left(\frac{\tau^*}{\text{SE}} - z_{1-\alpha/2}\right) + \Phi\left(-\frac{\tau^*}{\text{SE}} - z_{1-\alpha/2}\right)$$

## Sample Size Formula {visibility="uncounted"}

Rearranging the MDE formula to solve for $n$:

$$n = \frac{4\sigma^2(z_{1-\alpha/2} + z_{1-\beta})^2}{\text{MDE}^2}$$

**Example:** To detect MDE = 0.5 with $\sigma = 1.2$:

$$n = \frac{4(1.2)^2(1.96 + 0.84)^2}{(0.5)^2} = \frac{4 \times 1.44 \times 7.84}{0.25} \approx 181$$

(Per arm, so $N \approx 362$ total)
