---
title: "Session 1.3: Optimal Experimental Design & MIDA Framework"
subtitle: "Unit 1: Experimental Design"
author: "Sean Sylvia"
date: "January 28, 2026"
date-format: long
format:
  revealjs:
    theme: [default, ../style/gillings.scss]
    slide-number: true
    chalkboard: true
    preview-links: auto
    logo: ../images/logo.png
    footer: "HPM 883 | Session 1.3"
    transition: fade
    incremental: false
    smaller: true
    scrollable: true
    toc: true
    toc-depth: 1
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

## Today's Agenda {.center}

1.  The MIDA Framework
2.  Advanced Randomization (Blocking, Clustering)
3.  Variance Reduction (CUPED, ANCOVA)
4.  Design Diagnosis with DeclareDesign

::: notes
This session introduces the formal framework for declaring and diagnosing research designs.
:::

------------------------------------------------------------------------

## Introduction to DeclareDesign

### A Grammar for Research Design

DeclareDesign provides a unified framework:

-   **Model (M)**: Data generating process
-   **Inquiry (I)**: What we want to learn
-   **Data Strategy (D)**: How we collect data
-   **Answer Strategy (A)**: How we estimate

``` r
design <- declare_model(...) +
          declare_inquiry(...) +
          declare_sampling(...) +
          declare_assignment(...) +
          declare_estimator(...)
```

------------------------------------------------------------------------

## The MIDA Framework

```{mermaid}
%%| fig-alt: "MIDA framework showing Model, Inquiry, Data Strategy, and Answer Strategy as connected components"
flowchart LR
    M["Model<br>(DGP)"] --> I["Inquiry<br>(Estimand)"]
    I --> D["Data Strategy<br>(Sampling, Assignment)"]
    D --> A["Answer Strategy<br>(Estimator)"]
    A --> |"Diagnosis"| M
```

::: fragment
### Key Insight

Design declaration **before** data collection forces clear thinking about:

-   What we're trying to learn (estimand)
-   What assumptions we're making (model)
-   How we'll analyze the data (estimator)
:::

------------------------------------------------------------------------

## Example: Simple RCT

```{r}
#| echo: true
#| code-fold: false

library(DeclareDesign)
library(tidyverse)

# Declare the design
simple_rct <- declare_model(
  N = 100,
  U = rnorm(N),
  potential_outcomes(Y ~ 0.5 * Z + U)
) +
declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
declare_assignment(Z = complete_ra(N)) +
declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
declare_estimator(Y ~ Z, .method = lm_robust)
```

------------------------------------------------------------------------

## Running the Design

```{r}
#| echo: true

# Run once
run_design(simple_rct) |>
  select(inquiry, estimand, estimate, std.error) |>
  head(1)
```

::: fragment
### What just happened?

1.  Generated 100 units with potential outcomes
2.  Randomly assigned to treatment
3.  Revealed observed outcomes
4.  Estimated ATE with robust SEs
:::

------------------------------------------------------------------------

## Diagnosing the Design

```{r}
#| echo: true
#| cache: true

# Simulate 500 runs
diagnosis <- diagnose_design(simple_rct, sims = 500)
```

```{r}
#| echo: true
diagnosis |>
  reshape_diagnosis() |>
  select(Bias, RMSE, Power, Coverage)
```

------------------------------------------------------------------------

## Interpretation

| Metric       | Value  | Target |
|--------------|--------|--------|
| **Bias**     | \~0    | Near 0 |
| **RMSE**     | \~0.20 | Low    |
| **Power**    | \~0.60 | ≥ 0.80 |
| **Coverage** | \~0.95 | 0.95   |

::: fragment
### Problem: Underpowered!

With N=100 and true effect of 0.5, we only have \~60% power.

**Solutions:** - Increase sample size - Block on predictive covariates - Use CUPED/regression adjustment
:::

------------------------------------------------------------------------

## Your Turn: Modify the Design

Try changing:

1.  `N = 200` — What happens to power?
2.  `potential_outcomes(Y ~ 0.3 * Z + U)` — Smaller effect?
3.  `complete_ra(N, prob = 0.3)` — Unequal allocation?

``` r
# Template
my_design <- declare_model(
  N = ???,
  U = rnorm(N),
  potential_outcomes(Y ~ ??? * Z + U)
) +
declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
declare_assignment(Z = complete_ra(N, prob = ???)) +
declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
declare_estimator(Y ~ Z, .method = lm_robust)
```

------------------------------------------------------------------------

## Key Takeaways

1.  **MIDA** provides a formal grammar for research design

2.  **DeclareDesign** lets us simulate designs *before* collecting data

3.  **Diagnose before you analyze** — Power, bias, coverage

4.  **Design choices matter** — Blocking, sample size, allocation ratio

------------------------------------------------------------------------

## For Next Time

### Lab 1: Power Analysis by Simulation

-   Due: Sunday, February 2
-   Design and diagnose an RCT using DeclareDesign
-   Apply variance reduction techniques

**Readings:**

-   DeclareDesign documentation
-   Duflo, Glennerster, & Kremer (2007) — Design sections

------------------------------------------------------------------------

## Questions? {.center}

Office Hours: Tuesday 1-2pm, Wednesday 3-4pm

Canvas Discussion: For all course questions

::: notes
Open floor for questions about MIDA framework or DeclareDesign.
:::
