---
title: "Simulating Experimental Data with Noncompliance"
author: "Your Name"
date: "`r Sys.Date()`"
format: html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Simulating Experimental Data for a Randomized Experiment

In this example, we simulate data for a simple randomized experiment with one treatment and one control group. The randomization is at the individual level. Additionally, we introduce noncompliance, where some individuals assigned to the treatment group do not take the treatment.

------------------------------------------------------------------------

## Step 1: Setup and Parameters

Define the parameters for the experiment, such as the number of participants, compliance rate, and treatment effects.

```{r}
# Set seed for reproducibility
set.seed(123)

# Parameters
n <- 500  # Total number of participants
prob_treatment <- 0.5  # Probability of being assigned to treatment
compliance_rate <- 0.8  # Proportion of treated participants who comply

# Treatment effect
baseline_mean <- 50  # Mean outcome for control group
treatment_effect <- 10  # Additional effect for treated participants
```

------------------------------------------------------------------------

## Step 2: Random Assignment and Compliance

Simulate random assignment to treatment and compliance behavior.

```{r}
# Simulate random assignment
assignment <- rbinom(n, 1, prob_treatment)  # 1 = assigned to treatment, 0 = control

# Simulate compliance behavior
# If assigned to treatment, comply with probability `compliance_rate`
compliance <- ifelse(assignment == 1, rbinom(n, 1, compliance_rate), 0)  # 1 = complied, 0 = did not comply
```

------------------------------------------------------------------------

## Step 3: Simulate Outcomes

Simulate the outcome variable based on compliance and treatment assignment.

```{r}
# Simulate outcomes
outcome <- baseline_mean + 
  (assignment * compliance * treatment_effect) +  # Effect for those who comply
  rnorm(n, mean = 0, sd = 5)  # Add random noise

# Combine data into a data frame
experiment_data <- data.frame(
  ID = 1:n,
  Assignment = assignment,
  Compliance = compliance,
  Outcome = outcome
)

# View the first few rows of the data
head(experiment_data)
```

------------------------------------------------------------------------

## Step 4: Export Data to CSV

Export the simulated data to a CSV file for further analysis.

```{r}
# Export the data to a CSV file
write.csv(experiment_data, "simulated_experiment_data.csv", row.names = FALSE)

# Confirm export
cat("Data has been exported to 'simulated_experiment_data.csv'")
```