---
title: "PS 1: Experimental Design — Solutions"
subtitle: "Solutions"
author: "HPM 883 Teaching Team"
date: today
draft: false
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
execute:
  echo: true
  eval: true
  warning: false
  cache: true
---

---

## Setup

```{r}
#| label: setup
#| cache: false

library(DeclareDesign)
library(fabricatr)
library(randomizr)
library(estimatr)
library(tidyverse)
library(knitr)

# Load clinic data
clinic_data <- read.csv("../data/ps-1-clinic-data.csv")

# Parameters from problem set
tau <- -0.5       # True ATE
icc <- 0.15       # Assumed ICC for base designs
sigma_w <- 1.0    # Within-clinic SD
sigma_b <- sqrt(icc / (1 - icc))  # Between-clinic SD (~0.420 for ICC=0.15)

set.seed(883001)
```

---

## Task 1: Conceptual Foundation (20 points)

*No code required. Grade based on written explanations.*

### 1.1 The Design Effect Problem (8 points)

**a) Define the ICC (3 pts)**

The **intraclass correlation coefficient (ICC)** measures the proportion of total outcome variance attributable to between-cluster differences:

$$\rho = \frac{\sigma_b^2}{\sigma_b^2 + \sigma_w^2}$$

where $\sigma_b^2$ is the between-cluster variance and $\sigma_w^2$ is the within-cluster variance. ICC ranges from 0 (no clustering) to 1 (all variance between clusters).

**b) Design Effect Formula (3 pts)**

The **design effect** (DEFF) for a cluster-randomized trial is:

$$DE = 1 + (m - 1)\rho$$

where $m$ is the number of individuals per cluster and $\rho$ is the ICC. The **effective sample size** is:

$$N_{eff} = \frac{N}{DE} = \frac{K \cdot m}{1 + (m-1)\rho}$$

As ICC increases, DE increases and effective sample size decreases. A clustered sample of $N$ individuals provides less information than $N$ independent individuals.

**c) Calculation (2 pts)**

With ICC = 0.15 and $m$ = 20 patients per clinic:

$$DE = 1 + (20 - 1)(0.15) = 1 + 2.85 = 3.85$$

With 20 clinics × 20 patients = 400 total:

$$N_{eff} = \frac{400}{3.85} \approx 104$$

Our 400-patient clustered sample provides only as much information as ~104 independent observations.

### 1.2 Variance Reduction Strategies (6 points)

**d) Lin (2013) Estimator (3 pts)**

The **Lin estimator** regresses the outcome on treatment, centered covariates, and treatment × centered-covariate interactions:

$$Y_i = \alpha + \tau Z_i + \beta(X_i - \bar{X}) + \gamma Z_i(X_i - \bar{X}) + \varepsilon_i$$

Unlike standard ANCOVA (which omits the interaction term), this specification:

1. **Centers covariates** so $\tau$ is estimated at the sample mean of $X$
2. **Includes the interaction** $Z_i(X_i - \bar{X})$, allowing treatment effects to vary with the covariate
3. **Guarantees** that the variance of $\hat{\tau}$ is no larger than the unadjusted estimator, regardless of the true data-generating process

Standard ANCOVA can actually *increase* variance if the covariate-outcome relationship differs between treatment groups. Lin's approach avoids this by saturating the model within each arm.

**e) CUPED (3 pts)**

**CUPED** (Controlled Experiments Using Pre-Experiment Data) uses pre-treatment measurements of the outcome (or strongly correlated variables) as covariates for variance reduction:

$$\hat{\tau}_{CUPED} = \hat{\tau} - \theta(\bar{X}_T - \bar{X}_C)$$

where $X$ is the pre-experiment measurement and $\theta$ is chosen to minimize variance.

CUPED provides the **largest variance reduction** when the pre-experiment measure is highly correlated with the post-experiment outcome (high $R^2$ between pre and post). In our context, baseline HbA1c is likely strongly correlated with follow-up HbA1c, making CUPED particularly effective.

CUPED is closely related to Lin's estimator — both leverage baseline covariates for variance reduction. CUPED emphasizes using *pre-experiment versions of the outcome*, while Lin provides a general regression framework.

### 1.3 Optimal Allocation (6 points)

**f) Derivation (6 pts)**

The variance of the treatment effect estimator in a CRT is:

$$\text{Var}(\hat{\tau}) = \frac{4}{K}\left(\sigma_b^2 + \frac{\sigma_w^2}{m}\right)$$

Subject to budget constraint: $B = c_c \cdot K + c_p \cdot m \cdot K = K(c_c + c_p \cdot m)$

Solving for $K$: $K = \frac{B}{c_c + c_p \cdot m}$

Substituting:

$$\text{Var}(\hat{\tau}) = \frac{4(c_c + c_p m)}{B}\left(\sigma_b^2 + \frac{\sigma_w^2}{m}\right)$$

Taking derivative with respect to $m$ and setting to zero:

$$\frac{d}{dm}: \quad c_p \sigma_b^2 - \frac{c_c \sigma_w^2}{m^2} = 0$$

Solving:

$$\boxed{m^* = \sqrt{\frac{c_c}{c_p} \cdot \frac{\sigma_w^2}{\sigma_b^2}} = \sqrt{\frac{c_c}{c_p} \cdot \frac{1-\rho}{\rho}}}$$

**Dependence on ICC**: As $\rho$ increases, $(1-\rho)/\rho$ decreases, so $m^*$ decreases. Higher ICC means more cluster-level variance, so the optimal design favors **more clusters with fewer patients each**.

With our parameters ($c_c = 2000$, $c_p = 500$, $\rho = 0.15$):

$$m^* = \sqrt{\frac{2000}{500} \cdot \frac{0.85}{0.15}} = \sqrt{4 \times 5.67} = \sqrt{22.67} \approx 4.8$$

So the optimal allocation is approximately **5 patients per clinic**, with $K = 150{,}000 / (2000 + 500 \times 5) = 33$ clinics.

---

## Task 2: Design Declaration and Diagnosis (30 points)

### 2.1 Simple Cluster RCT (10 points)

```{r}
#| label: design-simple

# Task 2.1: Simple Cluster RCT
# K=20 clinics, m=20 patients, ICC=0.15, difference-in-means

design_simple <-
  declare_model(
    clinics = add_level(N = 20,
      u_j = rnorm(N, 0, sigma_b),
      urban = rbinom(N, 1, 0.6)
    ),
    patients = add_level(N = 20,
      u_ij = rnorm(N, 0, sigma_w),
      baseline_hba1c = 8.5 + 0.4 * u_j + 0.5 * u_ij + rnorm(N, 0, 0.6),
      Y_Z_0 = 8.5 + u_j + u_ij,
      Y_Z_1 = Y_Z_0 + tau
    )
  ) +
  declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  declare_assignment(Z = cluster_ra(clusters = clinics)) +
  declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  declare_estimator(
    Y ~ Z, .method = lm_robust, clusters = clinics,
    inquiry = "ATE", label = "DIM"
  )
```

```{r}
#| label: diagnose-simple

diag_simple <- diagnose_design(design_simple, sims = 500)
diag_simple
```

### 2.2 Blocked Cluster RCT (10 points)

```{r}
#| label: design-blocked

# Task 2.2: Blocked by urban/rural, block fixed effects in estimator

design_blocked <-
  declare_model(
    clinics = add_level(N = 20,
      u_j = rnorm(N, 0, sigma_b),
      urban = rbinom(N, 1, 0.6)
    ),
    patients = add_level(N = 20,
      u_ij = rnorm(N, 0, sigma_w),
      baseline_hba1c = 8.5 + 0.4 * u_j + 0.5 * u_ij + rnorm(N, 0, 0.6),
      Y_Z_0 = 8.5 + u_j + u_ij,
      Y_Z_1 = Y_Z_0 + tau
    )
  ) +
  declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  declare_assignment(Z = block_and_cluster_ra(blocks = urban, clusters = clinics)) +
  declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  declare_estimator(
    Y ~ Z + urban, .method = lm_robust, clusters = clinics,
    inquiry = "ATE", label = "Blocked"
  )
```

```{r}
#| label: diagnose-blocked

diag_blocked <- diagnose_design(design_blocked, sims = 500)
diag_blocked
```

### 2.3 Covariate-Adjusted Design (10 points)

```{r}
#| label: design-lin

# Task 2.3: Lin estimator with baseline HbA1c adjustment

design_lin <-
  declare_model(
    clinics = add_level(N = 20,
      u_j = rnorm(N, 0, sigma_b),
      urban = rbinom(N, 1, 0.6)
    ),
    patients = add_level(N = 20,
      u_ij = rnorm(N, 0, sigma_w),
      baseline_hba1c = 8.5 + 0.4 * u_j + 0.5 * u_ij + rnorm(N, 0, 0.6),
      Y_Z_0 = 8.5 + u_j + u_ij,
      Y_Z_1 = Y_Z_0 + tau
    )
  ) +
  declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  declare_assignment(Z = cluster_ra(clusters = clinics)) +
  declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  declare_estimator(
    Y ~ Z, covariates = ~ baseline_hba1c, .method = lm_lin,
    clusters = clinics, inquiry = "ATE", label = "Lin"
  )
```

```{r}
#| label: diagnose-lin

diag_lin <- diagnose_design(design_lin, sims = 500)
diag_lin
```

### Task 2 Comparison

```{r}
#| label: task2-comparison

# Extract diagnosands from each design
extract_diag <- function(diag, design_name) {
  diag$diagnosands_df |>
    mutate(design = design_name)
}

comparison <- bind_rows(
  extract_diag(diag_simple, "2.1 Simple CRT"),
  extract_diag(diag_blocked, "2.2 Blocked CRT"),
  extract_diag(diag_lin, "2.3 Lin-adjusted")
)

comparison |>
  select(design, power, bias, rmse, coverage) |>
  kable(digits = 3, caption = "Design Comparison: Key Diagnosands")
```

---

## Task 3: Design Diagnostics (20 points)

For sensitivity analyses, we define a helper function for efficient power simulation across parameter grids. This is equivalent to `diagnose_design()` but faster for grid searches.

```{r}
#| label: sim-power-function
#| cache: false

sim_power <- function(K, m, icc_val, tau = -0.5, sims = 500,
                      use_lin = FALSE, seed = 883) {
  set.seed(seed)
  sb <- sqrt(icc_val / (1 - icc_val))

  reject <- logical(sims)
  for (s in 1:sims) {
    clinic_id <- rep(1:K, each = m)
    u_j <- rep(rnorm(K, 0, sb), each = m)
    u_ij <- rnorm(K * m, 0, 1)
    Z <- rep(sample(rep(0:1, length.out = K)), each = m)

    baseline <- 8.5 + 0.4 * u_j + 0.5 * u_ij + rnorm(K * m, 0, 0.6)
    Y <- 8.5 + u_j + u_ij + tau * Z

    df <- data.frame(Y = Y, Z = Z, clinic_id = clinic_id,
                     baseline = baseline)

    if (use_lin) {
      fit <- lm_lin(Y ~ Z, covariates = ~ baseline,
                    data = df, clusters = clinic_id)
    } else {
      fit <- lm_robust(Y ~ Z, data = df, clusters = clinic_id)
    }

    pval <- summary(fit)$coefficients["Z", "Pr(>|t|)"]
    reject[s] <- pval < 0.05
  }
  mean(reject)
}
```

### 3.1 ICC Sensitivity Analysis (10 points)

```{r}
#| label: icc-sensitivity

icc_grid <- seq(0.05, 0.30, by = 0.025)

icc_results <- map_dfr(icc_grid, function(icc_val) {
  tibble(
    icc = icc_val,
    Unadjusted = sim_power(K = 20, m = 20, icc_val = icc_val,
                           use_lin = FALSE),
    `Lin-adjusted` = sim_power(K = 20, m = 20, icc_val = icc_val,
                               use_lin = TRUE)
  )
}) |>
  pivot_longer(cols = c(Unadjusted, `Lin-adjusted`),
               names_to = "Estimator", values_to = "Power")
```

```{r}
#| label: icc-plot
#| fig-cap: "Power vs. ICC for Unadjusted and Lin-adjusted Estimators"
#| fig-width: 8
#| fig-height: 5

ggplot(icc_results, aes(x = icc, y = Power, color = Estimator)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "gray40") +
  annotate("text", x = 0.28, y = 0.82, label = "80% power",
           color = "gray40", size = 3.5) +
  scale_y_continuous(limits = c(0.3, 1), labels = scales::percent) +
  scale_color_manual(values = c("Unadjusted" = "#E74C3C",
                                "Lin-adjusted" = "#2980B9")) +
  labs(
    title = "Power Decreases as ICC Increases",
    subtitle = "20 clinics, 20 patients/clinic, ATE = -0.5",
    x = "Intraclass Correlation (ICC)",
    y = "Statistical Power"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

**Key findings:**

- **Unadjusted**: Power drops below 80% around ICC ≈ 0.08-0.10. By ICC = 0.15, power is ~65-75%.
- **Lin-adjusted**: The threshold is higher, around ICC ≈ 0.15-0.18. Covariate adjustment buys meaningful headroom.
- The gap between curves shows the value of covariate adjustment, which is consistent regardless of ICC.

### 3.2 Sample Size Requirements (10 points)

```{r}
#| label: sample-size-sims

# Vary number of clinics (fix m = 20 patients)
k_grid <- seq(10, 50, by = 5)
k_results <- map_dfr(k_grid, function(K_val) {
  tibble(
    K = K_val, m = 20,
    Power = sim_power(K = K_val, m = 20, icc_val = 0.15, use_lin = TRUE)
  )
})

# Vary patients per clinic (fix K = 20 clinics)
m_grid <- seq(10, 50, by = 5)
m_results <- map_dfr(m_grid, function(m_val) {
  tibble(
    K = 20, m = m_val,
    Power = sim_power(K = 20, m = m_val, icc_val = 0.15, use_lin = TRUE)
  )
})
```

```{r}
#| label: sample-size-plots
#| fig-cap: "Power vs. Number of Clinics (left) and Patients per Clinic (right)"
#| fig-width: 10
#| fig-height: 5

p1 <- ggplot(k_results, aes(x = K, y = Power)) +
  geom_line(linewidth = 1, color = "#2980B9") +
  geom_point(size = 2, color = "#2980B9") +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "gray40") +
  scale_y_continuous(limits = c(0.3, 1), labels = scales::percent) +
  labs(
    title = "More Clinics: Strong Power Gains",
    subtitle = "Fix 20 patients/clinic, ICC = 0.15, Lin estimator",
    x = "Number of Clinics (K)", y = "Power"
  ) +
  theme_minimal()

p2 <- ggplot(m_results, aes(x = m, y = Power)) +
  geom_line(linewidth = 1, color = "#27AE60") +
  geom_point(size = 2, color = "#27AE60") +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "gray40") +
  scale_y_continuous(limits = c(0.3, 1), labels = scales::percent) +
  labs(
    title = "More Patients: Diminishing Returns",
    subtitle = "Fix 20 clinics, ICC = 0.15, Lin estimator",
    x = "Patients per Clinic (m)", y = "Power"
  ) +
  theme_minimal()

# Side-by-side display
library(patchwork)
p1 + p2
```

**Key findings:**

- **More clinics > more patients per clinic**. Increasing K from 10 to 30 has a much larger effect on power than increasing m from 10 to 30.
- The minimum number of clinics for 80% power (with Lin adjustment, m=20) is approximately **25-30 clinics**.
- Adding patients beyond ~20 per clinic shows diminishing returns because the cluster-level variance ($\sigma_b^2$) is the bottleneck.
- This directly motivates the budget optimization in Task 4.

---

## Task 4: Budget-Constrained Optimization (15 points)

### 4.1 Feasible Designs (5 points)

```{r}
#| label: feasible-designs

budget <- 150000
c_c <- 2000   # clinic startup cost
c_p <- 500    # per-patient cost

# Analytical power function (faster than simulation for grid search)
analytical_power <- function(K, m, icc_val, tau = -0.5, alpha = 0.05) {
  sigma_b2 <- icc_val / (1 - icc_val)
  sigma_w2 <- 1.0
  var_tau <- (4 / K) * (sigma_b2 + sigma_w2 / m)
  se <- sqrt(var_tau)
  ncp <- abs(tau) / se
  df <- K - 2
  crit <- qt(1 - alpha / 2, df)
  1 - pt(crit, df, ncp) + pt(-crit, df, ncp)
}

# Enumerate all feasible integer (K, m) pairs
feasible <- expand.grid(K = 5:60, m = 2:30) |>
  filter(K * (c_c + c_p * m) <= budget) |>
  mutate(
    cost = K * (c_c + c_p * m),
    total_N = K * m,
    power = mapply(analytical_power, K, m, icc_val = 0.15)
  )

# Top 10 designs by power
feasible |>
  arrange(desc(power)) |>
  head(10) |>
  mutate(power = scales::percent(power, accuracy = 0.1)) |>
  kable(caption = "Top 10 Feasible Designs by Analytical Power (ICC = 0.15)")
```

### 4.2 Optimal Allocation (5 points)

```{r}
#| label: budget-heatmap
#| fig-cap: "Power Across Feasible Budget-Constrained Designs"
#| fig-width: 9
#| fig-height: 7

ggplot(feasible, aes(x = K, y = m, fill = power)) +
  geom_tile() +
  scale_fill_viridis_c(
    name = "Power",
    labels = scales::percent,
    option = "plasma"
  ) +
  geom_contour(aes(z = power), breaks = c(0.5, 0.6, 0.7, 0.8),
               color = "white", linewidth = 0.5) +
  labs(
    title = "Power Heatmap: Budget-Constrained Design Space",
    subtitle = sprintf("Budget = $%s | Clinic cost = $%s | Patient cost = $%s",
                       scales::comma(budget), scales::comma(c_c), scales::comma(c_p)),
    x = "Number of Clinics (K)",
    y = "Patients per Clinic (m)"
  ) +
  theme_minimal()
```

```{r}
#| label: budget-optimal

# Identify optimal design
optimal <- feasible |> slice_max(power, n = 1, with_ties = FALSE)

cat("Optimal budget-constrained design (analytical):\n")
cat(sprintf("  K = %d clinics, m = %d patients/clinic\n", optimal$K, optimal$m))
cat(sprintf("  Total N = %d, Cost = $%s\n",
            optimal$total_N, scales::comma(optimal$cost)))
cat(sprintf("  Analytical power = %.1f%%\n", optimal$power * 100))
```

```{r}
#| label: validate-optimal

# Validate optimal with DeclareDesign simulation
opt_K <- optimal$K
opt_m <- optimal$m

design_optimal <-
  declare_model(
    clinics = add_level(N = opt_K,
      u_j = rnorm(N, 0, sigma_b)
    ),
    patients = add_level(N = opt_m,
      u_ij = rnorm(N, 0, sigma_w),
      baseline_hba1c = 8.5 + 0.4 * u_j + 0.5 * u_ij + rnorm(N, 0, 0.6),
      Y_Z_0 = 8.5 + u_j + u_ij,
      Y_Z_1 = Y_Z_0 + tau
    )
  ) +
  declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  declare_assignment(Z = cluster_ra(clusters = clinics)) +
  declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  declare_estimator(
    Y ~ Z, .method = lm_robust, clusters = clinics,
    inquiry = "ATE", label = "Unadjusted"
  ) +
  declare_estimator(
    Y ~ Z, covariates = ~ baseline_hba1c, .method = lm_lin,
    clusters = clinics, inquiry = "ATE", label = "Lin-adjusted"
  )

diag_optimal <- diagnose_design(design_optimal, sims = 500)
diag_optimal
```

### 4.3 Analytical vs. Simulation Comparison (5 points)

```{r}
#| label: analytical-comparison

# Analytical optimal from Task 1.3f
m_opt_analytical <- sqrt((c_c / c_p) * (1 - icc) / icc)
K_opt_analytical <- budget / (c_c + c_p * round(m_opt_analytical))

cat("Analytical optimal allocation (Task 1.3f formula):\n")
cat(sprintf("  m* = sqrt(%d/%d * %.2f/%.2f) = sqrt(%.2f) = %.2f ≈ %d\n",
            c_c, c_p, 1 - icc, icc,
            (c_c / c_p) * (1 - icc) / icc,
            m_opt_analytical, round(m_opt_analytical)))
cat(sprintf("  K* = %s / (%d + %d × %d) = %.1f ≈ %d\n",
            scales::comma(budget), c_c, c_p, round(m_opt_analytical),
            K_opt_analytical, floor(K_opt_analytical)))
cat(sprintf("\nSimulation-based optimal: K = %d, m = %d\n", opt_K, opt_m))
cat(sprintf("Analytical optimal:       K = %d, m = %d\n",
            floor(K_opt_analytical), round(m_opt_analytical)))
```

**Discussion:**

The analytical formula gives $m^* \approx 5$ and $K^* \approx 33$, which closely matches the simulation-based optimum. Small discrepancies arise because:

1. **Integer constraints**: The formula gives non-integer $m^*$; rounding changes $K$ slightly
2. **Small-sample effects**: The analytical formula uses a z-test approximation, while simulations use t-tests with cluster-robust SEs (which account for finite K)
3. **Estimator choice**: The analytical formula assumes a simple difference-in-means; the Lin estimator shifts the optimal slightly

The close agreement validates both the analytical formula and the simulation approach.

---

## Task 5: Executive Summary (15 points)

::: {.callout-note}
## Example Response

This is a sample response at the level expected for full credit. Your own framing may differ.
:::

---

**MEMORANDUM**

**To:** CEO Barnaby Beta and the Board of Directors, St. Null's Memorial Hospital

**From:** Biostatistics Consulting Team

**Re:** Recommended Study Design for the Integrated Care Coordination Program (ICCP) Evaluation

**Date:** February 2026

---

**The Problem with the Current Plan**

Dr. P-Hacker's proposed study design would randomize 400 individual patients and projects 90% power to detect our target effect. However, this calculation contains a critical flaw: it treats each patient as an independent observation.

In reality, the ICCP will be implemented at the clinic level — entire clinics will either receive the program or continue standard care. Patients within the same clinic share common experiences (same care team, local demographics, clinic culture), which means they provide less independent information than 400 separate patients would. When we account for this clustering, our actual ability to detect the treatment effect drops from 90% to roughly **60-65%** — meaning we have only a coin-flip chance of detecting a real improvement, even if the program works.

**Our Recommended Design**

After systematic analysis of hundreds of possible study configurations, we recommend:

| Parameter | Recommendation |
|-----------|---------------|
| Number of clinics | **33** (17 treatment, 16 control) |
| Patients per clinic | **5** |
| Total patients | 165 |
| Estimated budget | ~$148,500 (within the $150,000 limit) |
| Expected power | **75-85%** (with baseline HbA1c adjustment) |

The key insight is that **more clinics with fewer patients each** provides more statistical information than fewer clinics with many patients. This is because the clinic is our unit of comparison — adding a 34th clinic gives us a new independent data point, while adding a 21st patient to an existing clinic gives us redundant information.

We further recommend measuring each patient's **baseline HbA1c** before randomization and using a statistical technique called covariate adjustment. This leverages the strong correlation between baseline and follow-up HbA1c to sharpen our comparison, boosting power from ~62% to approximately 80%.

**Key Assumptions and Caveats**

Our recommendation depends on several assumptions:

- **Intraclass correlation (ICC) = 0.15**: If clinics are more similar internally (higher ICC), we would need even more clinics. Our sensitivity analysis shows that if ICC exceeds 0.20, power drops below 75% even with covariate adjustment.
- **Effect size = 0.5% HbA1c reduction**: If the true effect is smaller, more clinics would be needed.
- **Enrollment feasibility**: We assume 5 eligible patients per clinic can be recruited. If some clinics have fewer eligible patients, we should enroll additional clinics as backup.

**Bottom Line**

We strongly recommend proceeding with 33 clinics and 5 patients each, using baseline HbA1c measurement for covariate adjustment. This design maximizes our ability to detect a real improvement within the available budget. We advise against the original 20-clinic plan, which would be both over-budget and underpowered.

---

---

## Session Info

```{r}
#| label: session-info
#| cache: false

sessionInfo()
```

---

*Last updated: `r Sys.Date()`*

