---
title: "Code-Along X.Y: [TOPIC]"
subtitle: "Unit X: [UNIT TITLE]"
author: "Sean Sylvia"
date: "[DATE]"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Overview

**Learning Objective:** [What students will be able to do after completing this code-along]

**Method:** [Brief description of the method/technique being demonstrated]

**Packages Used:**

```{r}
#| label: setup
#| message: false

library(tidyverse)
# library([additional packages])

# Set seed for reproducibility
set.seed(42)
```

---

## 1. Concept Illustration

### What Does This Method Do?

[2-3 paragraphs explaining the concept, intuition, and when to use it]

### Key Assumptions

1. **[Assumption 1]:** [Description and why it matters]
2. **[Assumption 2]:** [Description and why it matters]
3. **[Assumption 3]:** [Description and why it matters]

### Visual Intuition

```{r}
#| label: concept-viz
#| fig-cap: "[Caption describing what this visualization shows]"
#| fig-alt: "[Alt text for accessibility]"

# Visualization code
```

---

## 2. Success Case: When It Works

### Setup: Ideal Conditions

[Description of the data generating process where all assumptions are satisfied]

```{r}
#| label: success-dgp

# Generate data where assumptions hold
n <- 1000

# Data generating process
# df_success <- tibble(
#   x = ...,
#   y = ...
# )
```

### Apply the Method

```{r}
#| label: success-analysis

# Apply method to ideal data
# result_success <- method(...)
```

### Verify It Works

```{r}
#| label: success-verify

# Show that method recovers true parameters
# Compare estimate to truth
```

::: {.callout-tip}
## Why It Works
[Explain why the method succeeds under these conditions]
:::

---

## 3. Failure Case: When It Breaks

### Setup: Assumption Violation

[Description of how we're violating one or more key assumptions]

```{r}
#| label: failure-dgp

# Generate data where assumptions are violated
# df_failure <- tibble(
#   x = ...,
#   y = ...
# )
```

### Apply the Method Anyway

```{r}
#| label: failure-analysis

# Apply same method to problematic data
# result_failure <- method(...)
```

### See the Breakdown

```{r}
#| label: failure-verify

# Show that method fails to recover true parameters
# Visualize the bias or other problems
```

::: {.callout-warning}
## Why It Fails
[Explain the specific assumption violation and its consequences]
:::

### Diagnosing the Problem

```{r}
#| label: diagnosis

# Diagnostic checks that would reveal the problem
```

---

## 4. Step-by-Step Analysis

Now let's work through a complete analysis from start to finish.

### 4.1 Data Preparation

```{r}
#| label: step-data

# Load/prepare data
# data <- read_csv("data/example.csv")
```

### 4.2 Exploratory Analysis

```{r}
#| label: step-explore

# Summary statistics
# Visualizations
```

### 4.3 Model Specification

```{r}
#| label: step-model

# Specify and fit model
```

### 4.4 Results

```{r}
#| label: step-results

# Extract and display results
```
### 4.5 Interpretation

[Guide students through interpreting the output]

- **Point estimate:** [Interpretation]
- **Confidence interval:** [Interpretation]
- **Practical significance:** [Interpretation]

---

## 5. Exercises

### Exercise 1: [TITLE]

[Exercise description]

```{r}
#| label: exercise-1
#| eval: false

# Starter code
# YOUR CODE HERE
```

::: {.callout-note collapse="true"}
## Hint
[Hint for students who are stuck]
:::

### Exercise 2: [TITLE]

[Exercise description]

```{r}
#| label: exercise-2
#| eval: false

# Starter code
# YOUR CODE HERE
```

---

## 6. AI Guidance Notes

::: {.callout-important}
## Using AI Assistants for This Topic
:::

### When AI Helps

- [Scenario where AI assistance is valuable]
- [Type of question to ask AI]

### When to Be Careful

- [Common AI mistake or hallucination for this topic]
- [Signs that AI output may be wrong]

### Verification Checklist

Before trusting AI-generated code for [TOPIC]:

- [ ] [Check 1]
- [ ] [Check 2]
- [ ] [Check 3]

### Example Prompt

> "[Example of a good prompt to use with an AI assistant for this topic]"

---

## Summary

**Key Takeaways:**

1. [Takeaway 1]
2. [Takeaway 2]
3. [Takeaway 3]

**Common Pitfalls:**

- [Pitfall 1]
- [Pitfall 2]

**Next Steps:** [What comes next in the course sequence]

---

## Resources

### Required Reading

- [Primary textbook reference]

### Further Reading

- [Additional resource]

### R Documentation

- `?[function_name]` — [Brief description]
