---
title: "Lab X: [LAB TITLE]"
subtitle: "Unit X: [UNIT TITLE]"
author: "[Student Name]"
date: "[DATE]"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

## Learning Objectives

By completing this lab, you will:

1. [Objective 1]
2. [Objective 2]
3. [Objective 3]

## Instructions

- **Due:** [DATE] by 11:59 PM
- **Submission:** Push to your GitHub Classroom repository
- **Collaboration:** [Policy — individual or group]
- **AI Use:** [Policy — permitted with documentation, etc.]

::: {.callout-important}
## Before You Begin
Ensure you have completed the setup verification and have all required packages installed.
:::

## Setup

```{r}
#| label: setup

library(tidyverse)
# Add other required packages

# Load data
# data <- read_csv("data/[filename].csv")
```

---

## Part 1: [SECTION TITLE] (XX points)

[Context setting — brief narrative if applicable]

### Task 1.1 (X points)

[Clear task description]

```{r}
#| label: task-1-1

# YOUR CODE HERE

```

**Your interpretation:** [Prompt for written response if needed]

> [Student writes here]

### Task 1.2 (X points)

[Clear task description]

```{r}
#| label: task-1-2

# YOUR CODE HERE

```

---

## Part 2: [SECTION TITLE] (XX points)

### Task 2.1 (X points)

[Clear task description]

```{r}
#| label: task-2-1

# YOUR CODE HERE

```

### Task 2.2 (X points)

[Clear task description]

```{r}
#| label: task-2-2

# YOUR CODE HERE

```

---

## Part 3: Conceptual Questions (XX points)

### Question 3.1 (X points)

[Conceptual question requiring understanding, not just computation]

> [Student writes here]

### Question 3.2 (X points)

[Conceptual question]

> [Student writes here]

---
## Part 4: Reflection (X points)

### What was the most challenging part of this lab?

> [Student writes here]

### What would you do differently if you were to repeat this analysis?

> [Student writes here]

---

## Submission Checklist

Before submitting, verify:

- [ ] All code chunks run without errors
- [ ] All written responses are complete
- [ ] Document renders to HTML successfully
- [ ] All files are committed and pushed to GitHub
- [ ] If AI was used, usage is documented below

## AI Usage Documentation

If you used AI assistance, document it here:

**AI Tool(s) Used:** [e.g., Claude, ChatGPT, GitHub Copilot]

**How AI Was Used:**

> [Describe how you used AI — e.g., "debugging Task 2.1", "explaining concept X"]

**Verification Steps Taken:**

> [How you verified AI output was correct]
