---
title: "Session X.Y: [SESSION TITLE]"
subtitle: "Unit X: [UNIT TITLE]"
author: "Sean Sylvia"
date: "[DATE]"
format:
  revealjs:
    theme: [default, ../style/custom.scss]
    slide-number: true
    chalkboard: true
    preview-links: auto
    logo: ../images/unc-logo.png
    footer: "HPM 883 | Spring 2026"
    transition: fade
    incremental: true
execute:
  echo: true
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(knitr)
library(kableExtra)

# Set default options
knitr::opts_chunk$set(
  fig.width = 10,
  fig.height = 6,
  fig.retina = 2,
  out.width = "100%"
)

# Custom theme for plots
theme_set(theme_minimal(base_size = 14))
```

# Learning Objectives

## By the end of this session, you will...

1. [Objective 1]
2. [Objective 2]
3. [Objective 3]

::: notes
Speaker notes for objectives slide.
Remind students how this connects to prior sessions.
:::

# Part 1: [SECTION TITLE]

## [Slide Title]

::: {.columns}
::: {.column width="50%"}
**Key Point:**

- Bullet 1
- Bullet 2
- Bullet 3
:::

::: {.column width="50%"}
```{r}
#| label: example-plot
#| echo: false
#| fig-alt: "[Alt text describing the plot for accessibility]"

# Example visualization
# ggplot(...) + ...
```
:::
:::

::: notes
Speaker notes explaining the concept in more depth.
Include examples, anecdotes, or questions to pose to class.
:::

## Key Concept: [CONCEPT NAME]

::: {.callout-note}
## Definition
[Formal definition of the concept]
:::

::: fragment
**Intuition:** [Plain language explanation]
:::

::: fragment
**Example:** [Concrete example from health services research]
:::

# Part 2: [SECTION TITLE]

## Mathematical Formulation

$$
[LaTeX formula]
$$

where:

- $X$ = [description]
- $Y$ = [description]

::: notes
Walk through each term. Connect to intuition from previous slide.
:::

## R Implementation

```{r}
#| label: code-example
#| eval: false

# Example code implementation
# result <- function_name(
#   data = df,
#   formula = y ~ x,
#   ...
# )
```

::: fragment
**Key parameters:**

- `param1`: [description]
- `param2`: [description]
:::

# Part 3: Application

## Example: [APPLIED EXAMPLE TITLE]

**Context:** [Brief description of health policy/HSR context]

**Question:** [Research question being addressed]

**Data:** [Description of data source]

## Results

```{r}
#| label: applied-example
#| echo: true

# Applied example code
```

## Interpretation

- **Finding 1:** [Interpretation]
- **Finding 2:** [Interpretation]

::: {.callout-warning}
## Caution
[Common misinterpretation or limitation to highlight]
:::

# Wrap-Up

## Key Takeaways

1. [Takeaway 1]
2. [Takeaway 2]
3. [Takeaway 3]

## For Next Time

**Readings:**

- [Reading assignment]

**Prepare:**

- [Preparation task, if any]

## Questions?

::: {.r-fit-text}
Office Hours: [Time/Location]
:::

# Appendix {visibility="uncounted"}

## Additional Resources

**Remedial:**

- [Resource for students needing background]

**Advanced:**

- [Resource for students wanting to go deeper]

## References

::: {#refs}
:::
