---
title: "Session 0.2: Reproducible Research & Computing Setup"
subtitle: "Unit 0: Foundations"
author: "Sean Sylvia"
date: "January 12, 2026"
date-format: long
format:
  revealjs:
    theme: [default, ../style/gillings.scss]
    slide-number: true
    chalkboard: true
    preview-links: auto
    logo: ../images/logo.png
    footer: "HPM 883 | Session 0.2"
    transition: fade
    incremental: false
    smaller: true
    scrollable: true
    toc: true
    toc-depth: 1
    code-fold: false
    highlight-style: github
execute:
  echo: true
  warning: false
  message: false
---

## Today's Agenda {.center}

1.  Reproducible research principles
2.  Computing environment setup
3.  Introduction to DeclareDesign
4.  Simulating your first RCT

::: notes
This is a hands-on session. Have your laptops ready. We'll be coding together.
:::

------------------------------------------------------------------------

## Why Reproducibility Matters

::::: columns
::: {.column width="50%"}
### The Reproducibility Crisis

-   70%+ of researchers have failed to reproduce another's work
-   50%+ have failed to reproduce their *own* work
-   Retractions increasing exponentially
:::

::: {.column width="50%"}
### Our Solution

-   Version control (Git)
-   Package management (renv)
-   Pipeline automation (targets)
-   Literate programming (Quarto)
:::
:::::

::: callout-important
## Course Requirement

All submitted work must be reproducible. If we can't run your code, we can't grade it.
:::

------------------------------------------------------------------------

## The Reproducibility Stack

```{mermaid}
%%| fig-alt: "Four layers of reproducibility: Quarto (reports) on top, targets (pipeline), renv (packages), Git (version control) at bottom"
flowchart TB
    subgraph L4["Literate Programming"]
        Q["Quarto: Code + Narrative"]
    end

    subgraph L3["Pipeline"]
        T["targets: Automated workflow"]
    end

    subgraph L2["Environment"]
        R["renv: Package versions"]
    end

    subgraph L1["Version Control"]
        G["Git: Track changes"]
    end

    L4 --> L3
    L3 --> L2
    L2 --> L1
```

------------------------------------------------------------------------

## Git & GitHub

### Why Version Control?

-   **Track changes**: See exactly what changed when
-   **Collaborate**: Multiple people, same codebase
-   **Backup**: Your work lives in the cloud
-   **Submission**: GitHub Classroom for assignments

::: fragment
### Basic Workflow

``` bash
git clone <repo-url>    # Copy repo to local
git add .               # Stage changes
git commit -m "message" # Save snapshot
git push                # Upload to GitHub
```
:::

------------------------------------------------------------------------

## Package Management with renv

### The Problem

``` r
# Works on your machine...
install.packages("dplyr")  # Version 1.1.4

# Breaks on mine...
# I have dplyr 0.8.3
```

::: fragment
### The Solution: renv

``` r
# First time setup (already done in templates)
renv::init()

# Restore packages (you run this)
renv::restore()

# After adding packages
renv::snapshot()
```
:::

------------------------------------------------------------------------

## Pipeline Automation with targets

### Why targets?

-   Only re-run what changed
-   Document dependencies automatically
-   Scale to complex analyses

::: fragment
### Basic Structure

``` r
# _targets.R
list(
  tar_target(raw_data, read_csv("data/raw.csv")),
  tar_target(clean_data, clean_fn(raw_data)),
  tar_target(model, fit_model(clean_data)),
  tar_target(results, summarize(model))
)
```
:::

::: callout-tip
We'll use targets for the capstone project. Labs use simpler structure.
:::

------------------------------------------------------------------------

## Literate Programming with Quarto

### Code + Narrative = Reproducible Report

``` markdown
## Methods

We estimate the treatment effect using:

`​``{r}
model <- lm_robust(outcome ~ treatment, data = df)
`​``

The estimated effect is 0.42.
```

::: fragment
**Output:** A document where code, results, and interpretation live together.
:::

------------------------------------------------------------------------

## Your Computing Environment

### Two-Tier Architecture

| Tier            | Platform                 | When to Use                    |
|-----------------|--------------------------|--------------------------------|
| **Primary**     | GitHub Codespaces        | Labs, code-alongs, assignments |
| **Alternative** | Local (Positron/RStudio) | Capstone, offline work         |

::: callout-note
## Zero Local Setup Required

GitHub Codespaces runs a full RStudio Server in your browser. No installation needed!
:::

------------------------------------------------------------------------

## Live Demo: GitHub Codespaces

Let's walk through:

1.  Accept assignment via GitHub Classroom link
2.  Click "Open in Codespace" → browser-based RStudio launches
3.  Run `renv::restore()` to install packages
4.  Work on your code
5.  Commit and push when done — no download needed

::: notes
Walk through the demo slowly. Pause for questions. This is where students often get stuck.
:::

------------------------------------------------------------------------

## Introduction to DeclareDesign

### A Grammar for Research Design

DeclareDesign provides a unified framework:

-   **Model (M)**: Data generating process
-   **Inquiry (I)**: What we want to learn
-   **Data Strategy (D)**: How we collect data
-   **Answer Strategy (A)**: How we estimate

``` r
design <- declare_model(...) +
          declare_inquiry(...) +
          declare_sampling(...) +
          declare_assignment(...) +
          declare_estimator(...)
```

------------------------------------------------------------------------

## The MIDA Framework

```{mermaid}
%%| fig-alt: "MIDA framework showing Model, Inquiry, Data Strategy, and Answer Strategy as connected components"
flowchart LR
    M["Model<br>(DGP)"] --> I["Inquiry<br>(Estimand)"]
    I --> D["Data Strategy<br>(Sampling, Assignment)"]
    D --> A["Answer Strategy<br>(Estimator)"]
    A --> |"Diagnosis"| M
```

::: fragment
### Key Insight

Design declaration **before** data collection forces clear thinking about:

-   What we're trying to learn (estimand)
-   What assumptions we're making (model)
-   How we'll analyze the data (estimator)
:::

------------------------------------------------------------------------

## Example: Simple RCT

```{r}
#| echo: true
#| code-fold: false

library(DeclareDesign)
library(tidyverse)

# Declare the design
simple_rct <- declare_model(
  N = 100,
  U = rnorm(N),
  potential_outcomes(Y ~ 0.5 * Z + U)
) +
declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
declare_assignment(Z = complete_ra(N)) +
declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
declare_estimator(Y ~ Z, .method = lm_robust)
```

------------------------------------------------------------------------

## Running the Design

```{r}
#| echo: true

# Run once
run_design(simple_rct) |>
  select(inquiry, estimand, estimate, std.error) |>
  head(1)
```

::: fragment
### What just happened?

1.  Generated 100 units with potential outcomes
2.  Randomly assigned to treatment
3.  Revealed observed outcomes
4.  Estimated ATE with robust SEs
:::

------------------------------------------------------------------------

## Diagnosing the Design

```{r}
#| echo: true
#| cache: true

# Simulate 500 runs
diagnosis <- diagnose_design(simple_rct, sims = 500)
```

```{r}
#| echo: true
diagnosis |>
  reshape_diagnosis() |>
  select(Bias, RMSE, Power, Coverage)
```

------------------------------------------------------------------------

## Interpretation

| Metric       | Value  | Target |
|--------------|--------|--------|
| **Bias**     | \~0    | Near 0 |
| **RMSE**     | \~0.20 | Low    |
| **Power**    | \~0.60 | ≥ 0.80 |
| **Coverage** | \~0.95 | 0.95   |

::: fragment
### Problem: Underpowered!

With N=100 and true effect of 0.5, we only have \~60% power.

**Solutions:** - Increase sample size - Block on predictive covariates - Use CUPED/regression adjustment
:::

------------------------------------------------------------------------

## Your Turn: Modify the Design

Try changing:

1.  `N = 200` — What happens to power?
2.  `potential_outcomes(Y ~ 0.3 * Z + U)` — Smaller effect?
3.  `complete_ra(N, prob = 0.3)` — Unequal allocation?

``` r
# Template
my_design <- declare_model(
  N = ???,
  U = rnorm(N),
  potential_outcomes(Y ~ ??? * Z + U)
) +
declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
declare_assignment(Z = complete_ra(N, prob = ???)) +
declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
declare_estimator(Y ~ Z, .method = lm_robust)
```

------------------------------------------------------------------------

## Getting Help

### When Stuck

1.  **Canvas Discussion Board** — Quick questions
2.  **Office Hours** — Tuesday 1-2pm, Wednesday 3-4pm
3.  **Email** — For private concerns

### Troubleshooting Guide

See [Computing Page](../computing.qmd#troubleshooting) for common issues.

::: callout-tip
## Pro Tip

Google error messages. Stack Overflow has solved most R problems.
:::

------------------------------------------------------------------------

## Key Takeaways

1.  **Reproducibility** is non-negotiable — renv + Git + Quarto

2.  **DeclareDesign** lets us simulate designs *before* collecting data

3.  **Diagnose before you analyze** — Power, bias, coverage

4.  **GitHub Codespaces** = zero local setup required

------------------------------------------------------------------------

## For Next Time

### Unit 1: Experimental Design

**Readings:**

-   Chernozhukov et al. Ch. 1-2 — Foundations of causal inference
-   Wager Ch. 1 — Potential outcomes framework

**Prepare:**

-   Review course computing setup if needed
-   Bring questions about Codespaces workflow

::: notes
Remind students to familiarize themselves with the GitHub workflow before Unit 1.
:::

------------------------------------------------------------------------

## Questions? {.center}

Office Hours: Tuesday 1-2pm, Wednesday 3-4pm

Canvas Discussion: For all course questions

::: notes
Open floor for questions. This is usually when students reveal confusion about setup.
:::