---
title: "Session 3.0: Causal Random Forests"
subtitle: "HPM 883: Advanced Quantitative Methods"
author: "Sean Sylvia"
date: "March 25, 2026"
date-format: long
format:
  revealjs:
    theme: [default, ../style/gillings.scss]
    slide-number: true
    transition: fade
    progress: true
    chalkboard: true
    smaller: true
    scrollable: true
    logo: ../images/logo.png
    footer: "HPM 883 | Session 3.0: Causal Random Forests"
    toc: true
    toc-depth: 1
    preview-links: auto
    incremental: true
    html-math-method: mathjax
execute:
  echo: true
  warning: false
  message: false
---

# Introduction

## The Shift

::: {.callout-note appearance="simple"}
## The Arc of the Course

**Unit 1 (Weeks 1-3):** Experimental Design — *"Does it work?"*

**Unit 2 (Weeks 4-10):** DML & Doubly-Robust Estimation — *"Does it work?"* (with ML)

**→ Unit 3 (Weeks 10-11):** Causal Forests & Meta-Learners — *"For whom does it work?"*

**Unit 4 (Week 11):** Policy Learning — *"What should we do?"*

**Unit 5 (Week 12):** Adaptive Experiments — *"What should we do?"* (dynamically)
:::

## Why "For Whom?"

Suppose a new health intervention reduces hospitalizations by 15% on average.

. . .

Should we roll it out to everyone?

. . .

Maybe. But what if:

- It reduces hospitalizations by **40%** for patients with multiple comorbidities
- It has **no effect** on young, healthy patients
- It actually **increases** costs for a subgroup due to side effects

. . .

The **average** hides crucial variation. Policy needs to know **who benefits**.

------------------------------------------------------------------------

## From ATE to CATE

**Average Treatment Effect (Unit 2):**

$$\tau = E[Y(1) - Y(0)]$$

. . .

**Conditional Average Treatment Effect (Unit 3):**

$$\tau(x) = E[Y(1) - Y(0) \mid X = x]$$

. . .

$\tau(x)$ is the treatment effect for individuals with characteristics $x$.

. . .

| Estimand | Question | Example |
|----------|----------|---------|
| ATE | Does it work on average? | "The program reduces ER visits by 2.3 per year" |
| CATE | Does it work for *this person*? | "For diabetic patients over 65, the reduction is 5.1" |

. . .

::: {.callout-tip appearance="simple"}
The ATE is a **summary** of CATEs: $\tau = E[\tau(X)]$
:::

------------------------------------------------------------------------

## The Subgroup Analysis Trap

The traditional approach to HTE: run regressions with interaction terms or split into subgroups.

. . .

**Why this fails:**

1. **Data dredging:** With 20 covariates, there are millions of possible subgroups. Some will show "significant" effects by chance.

. . .

2. **Multiple testing:** Standard CIs don't account for searching across subgroups. A "significant" subgroup effect may be noise.

. . .

3. **Curse of dimensionality:** Interactions between 3+ variables create tiny cells with unreliable estimates.

. . .

4. **Pre-specification paradox:** If you pre-specify subgroups, you miss unexpected heterogeneity. If you don't, you're data dredging.

. . .

::: {.callout-warning appearance="simple"}
## The Fundamental Problem
We need a principled way to discover *which* covariates drive treatment effect heterogeneity — without overfitting or p-hacking.
:::

------------------------------------------------------------------------

# Part 1: From Trees to Causal Trees {.center}

## Quick Review: Decision Trees (Session 2.1b)

::: {.nonincremental}
Recall from our trees lecture:

- Trees recursively **partition** the feature space into regions
- Each split maximizes some criterion (e.g., reduction in MSE)
- Predictions are **averages within leaves**
:::

. . .

For **prediction** trees, the split criterion minimizes prediction error:

$$\text{Split to minimize:} \quad \sum_{i \in \text{left}} (Y_i - \bar{Y}_L)^2 + \sum_{i \in \text{right}} (Y_i - \bar{Y}_R)^2$$

. . .

**Key idea for today:** What if, instead of predicting $Y$, we want to predict $\tau(X)$ — the *treatment effect*?

------------------------------------------------------------------------

## Causal Trees: The Intuition

**Goal:** Partition the covariate space so that leaves contain groups with *similar treatment effects*.

. . .

**Naive approach:** Build a tree where the split criterion maximizes the *difference in treatment effects* between child nodes.

. . .

$$\text{Split to maximize:} \quad (\hat\tau_L - \hat\tau_R)^2$$

where $\hat\tau_L = \bar{Y}_{L,\text{treated}} - \bar{Y}_{L,\text{control}}$

. . .

**Problem:** If we use the same data to:

1. **Choose** where to split (find the partition), and
2. **Estimate** the treatment effect within each leaf

. . .

...we get **overfitting**. The tree will find splits that maximize *apparent* heterogeneity, even if it's noise.

------------------------------------------------------------------------

## The Honesty Principle

**Athey & Imbens (2016)** introduced a solution: **honest estimation**.

. . .

Split your data into two halves:

```
┌─────────────────────────────────────────────┐
│                 Full Sample                  │
│                                             │
│  ┌──────────────────┐  ┌──────────────────┐ │
│  │  Splitting Half  │  │ Estimation Half  │ │
│  │                  │  │                  │ │
│  │  Used to build   │  │  Used to compute │ │
│  │  the tree        │  │  leaf estimates  │ │
│  │  structure       │  │  (τ̂ in each     │ │
│  │                  │  │   leaf)          │ │
│  └──────────────────┘  └──────────────────┘ │
└─────────────────────────────────────────────┘
```

. . .

1. **Splitting sample:** Determine tree structure (which variables to split on, where)
2. **Estimation sample:** Estimate $\hat\tau$ in each leaf using fresh data

. . .

::: {.callout-tip appearance="simple"}
## Why This Works
The estimation sample never influenced the splits. So the leaf estimates are **unbiased** — no overfitting to the tree structure.
:::

------------------------------------------------------------------------

## Honest Trees: Step by Step

**Step 1:** Randomly split data 50/50 into splitting ($\mathcal{S}$) and estimation ($\mathcal{E}$) samples.

. . .

**Step 2:** Using $\mathcal{S}$ only, grow a tree that maximizes treatment effect heterogeneity across leaves.

. . .

**Step 3:** Drop the leaf estimates from $\mathcal{S}$. Push $\mathcal{E}$ observations through the tree structure.

. . .

**Step 4:** In each leaf $\ell$, estimate:
$$\hat\tau_\ell = \frac{1}{n_{\ell,1}} \sum_{i \in \ell: D_i=1} Y_i - \frac{1}{n_{\ell,0}} \sum_{i \in \ell: D_i=0} Y_i$$

using only estimation-sample observations that fall in leaf $\ell$.

. . .

**Step 5:** For a new observation with $X = x$, find its leaf and report $\hat\tau(x) = \hat\tau_\ell$.

------------------------------------------------------------------------

# Part 2: Causal Random Forests {.center}

::: {.callout-note appearance="minimal"}
**Key Papers:** Athey, Tibshirani & Wager (2019); Wager & Athey (2018)
:::

## From One Tree to a Forest

A single honest causal tree is:

- **Noisy** — depends heavily on the random split
- **High variance** — small changes in data → very different tree
- **Coarse** — can only assign a few distinct $\hat\tau$ values (one per leaf)

. . .

**Solution:** Build a **forest** of many causal trees.

. . .

Each tree uses:

- A random subsample of the data
- A random subset of features at each split
- Honest splitting (separate estimation sample)

. . .

The forest prediction: **average across all trees**

$$\hat\tau(x) = \frac{1}{B} \sum_{b=1}^{B} \hat\tau_b(x)$$

. . .

This is the same idea as a random forest for prediction — but adapted for *treatment effects*.

------------------------------------------------------------------------

## What Makes Causal Forests Special?

Three key innovations beyond standard random forests:

. . .

### 1. Honesty

Every tree uses separate samples for splitting and estimation. This ensures valid inference (not just good prediction).

. . .

### 2. Local Centering (Orthogonalization)

Before building the forest, we "partial out" confounders:

$$\tilde{Y}_i = Y_i - \hat{m}(X_i), \quad \tilde{D}_i = D_i - \hat{e}(X_i)$$

where $\hat{m}(X) = E[Y|X]$ and $\hat{e}(X) = P(D=1|X)$.

. . .

**Why?** This is the same orthogonalization from DML! It removes the effect of confounders on both outcome and treatment, so the forest focuses on *treatment effect heterogeneity*.

. . .

### 3. Forest-Based Inference

The forest structure provides **asymptotically valid confidence intervals** for $\hat\tau(x)$ — no bootstrap needed.

------------------------------------------------------------------------

## Local Centering: The DML Connection

::: {.callout-tip appearance="simple"}
## Unit 2 → Unit 3 Bridge
Local centering in causal forests uses the **same doubly-robust scores** you learned in Unit 2.
:::

. . .

Recall the AIPW score for observation $i$:

$$\hat\phi_i = \hat\mu_1(X_i) - \hat\mu_0(X_i) + \frac{D_i(Y_i - \hat\mu_1(X_i))}{\hat\pi(X_i)} - \frac{(1-D_i)(Y_i - \hat\mu_0(X_i))}{1-\hat\pi(X_i)}$$

. . .

In Unit 2, we averaged these scores to get the ATE: $\hat\tau = \frac{1}{n}\sum_i \hat\phi_i$

. . .

**Causal forests do something different:** they estimate how $\hat\phi_i$ varies as a function of $X_i$.

. . .

Instead of one global average, we get a **local average** — the CATE:

$$\hat\tau(x) = \sum_{i=1}^n \alpha_i(x) \cdot \hat\phi_i$$

where $\alpha_i(x)$ are forest-derived **adaptive nearest-neighbor weights**.

------------------------------------------------------------------------

## How the Forest Creates Weights

For a new point $x$, the forest computes weights $\alpha_i(x)$:

. . .

1. For each tree $b$, find the leaf containing $x$
2. All training observations in that leaf are "neighbors" of $x$ in tree $b$
3. Each neighbor gets weight $1/n_\ell$ (where $n_\ell$ is the leaf size)
4. Average weights across all $B$ trees

. . .

$$\alpha_i(x) = \frac{1}{B} \sum_{b=1}^{B} \frac{\mathbb{1}(X_i \in L_b(x))}{|L_b(x)|}$$

. . .

**Intuition:** Observations that frequently end up in the same leaf as $x$ are "similar" — they get higher weight. The forest *adaptively* defines neighborhoods.

. . .

::: {.callout-tip appearance="simple"}
A causal forest is a **locally weighted doubly-robust estimator** where the weights are learned from the data.
:::

------------------------------------------------------------------------

# Part 3: The `grf` Package {.center}

## Causal Forests in R

The `grf` package (Generalized Random Forests) makes this easy:

```{r}
#| eval: false
#| echo: true

library(grf)

# Fit a causal forest
cf <- causal_forest(
  X = X,           # Covariate matrix (n × p)
  Y = Y,           # Outcome vector
  W = W,           # Treatment indicator (0/1)
  num.trees = 2000 # Number of trees
)

# Predict CATEs for all observations
tau_hat <- predict(cf)$predictions

# Get confidence intervals
ci <- predict(cf, estimate.variance = TRUE)
ci$lower <- ci$predictions - 1.96 * sqrt(ci$variance.estimates)
ci$upper <- ci$predictions + 1.96 * sqrt(ci$variance.estimates)
```

. . .

That's it. Under the hood, `grf` handles honesty, local centering, and forest-based inference automatically.

------------------------------------------------------------------------

## Key `grf` Functions

| Function | Purpose |
|----------|---------|
| `causal_forest()` | Fit a causal forest |
| `predict()` | Estimate CATEs (with optional variance) |
| `average_treatment_effect()` | Forest-weighted ATE (doubly robust) |
| `test_calibration()` | Test whether heterogeneity is real |
| `variable_importance()` | Which covariates drive HTE? |
| `best_linear_projection()` | Project CATEs onto selected covariates |

. . .

### Testing for Heterogeneity

```{r}
#| eval: false
#| echo: true

# Is there real heterogeneity?
test_calibration(cf)
# Reports: mean.forest.prediction (should ≈ ATE)
#          differential.forest.prediction (>0 if real HTE)
```

. . .

If the "differential" coefficient is significantly positive, the forest is detecting real treatment effect heterogeneity — not just noise.

------------------------------------------------------------------------

## Variable Importance

```{r}
#| eval: false
#| echo: true

# Which variables drive heterogeneity?
vi <- variable_importance(cf)
names(vi) <- colnames(X)
sort(vi, decreasing = TRUE)[1:5]
```

. . .

**Interpretation:** Variables that appear more often in splits are more important for explaining *treatment effect variation* (not outcome prediction).

. . .

::: {.callout-warning appearance="simple"}
## Caution
Variable importance tells you which variables the forest uses for splits, but it doesn't tell you the *direction* or *magnitude* of the effect. Use `best_linear_projection()` for interpretable summaries.
:::

------------------------------------------------------------------------

## Best Linear Projection

```{r}
#| eval: false
#| echo: true

# Project CATEs onto interpretable covariates
blp <- best_linear_projection(cf, A = X[, c("age", "comorbidity_count")])
print(blp)
```

. . .

This fits: $\hat\tau(x) \approx \beta_0 + \beta_1 \cdot \text{age} + \beta_2 \cdot \text{comorbidities}$

. . .

- Gives you interpretable coefficients with standard errors
- "Each additional comorbidity increases the treatment effect by 0.8 units (SE = 0.3)"
- More robust than pre-specified subgroup analysis

------------------------------------------------------------------------

# Part 4: Practical Considerations {.center}

## When Do Causal Forests Work Well?

**Good settings:**

- Moderate to large sample sizes ($n > 500$, ideally $> 1000$)
- Enough variation in treatment effects across $X$
- Good overlap (same as Unit 2 — positivity matters)
- Mix of continuous and categorical covariates

. . .

**Challenging settings:**

- Very small samples (trees need enough observations per leaf)
- Very high-dimensional $X$ with sparse signal
- When heterogeneity is along a single variable (simpler methods work)
- When treatment effects are near zero everywhere

. . .

::: {.callout-tip appearance="simple"}
## Rule of Thumb
If the ATE is near zero, be skeptical of large CATEs in subgroups — the forest may be finding noise.
:::

------------------------------------------------------------------------

## Inference and Interpretation

**What the confidence intervals cover:**

- The forest provides pointwise CIs for $\tau(x)$ at each $x$
- These are valid asymptotically under regularity conditions
- They account for the adaptive nature of the forest

. . .

**What they don't cover:**

- Simultaneous coverage across all $x$ (no uniform bands)
- Inference on which specific variables "matter" for HTE

. . .

**Best practices:**

1. **Always run `test_calibration()`** — check that heterogeneity is real
2. **Report the ATE** alongside CATEs — how much variation is there?
3. **Use `best_linear_projection()`** for interpretable HTE summaries
4. **Check overlap** — extreme propensity scores mean unreliable CATEs in those regions

------------------------------------------------------------------------

## Health Policy Application: Targeting

One of the most powerful applications of CATEs — **treatment targeting**:

. . .

1. Estimate $\hat\tau(x)$ for all individuals in your sample
2. Rank individuals by their predicted benefit
3. Under a **budget constraint**, treat those with the highest CATEs

. . .

**Example:** A community health worker intervention costs $500/patient.

- Average effect: 1.2 fewer ER visits ($\hat\tau = -1.2$)
- Top quartile by CATE: 3.8 fewer ER visits ($\hat\tau = -3.8$)
- Bottom quartile: 0.1 fewer ER visits ($\hat\tau = -0.1$)

. . .

**Targeting the top quartile** achieves 3× the impact per dollar.

. . .

::: {.callout-note appearance="simple"}
This is a preview of **Unit 4 (Policy Learning)** — we'll formalize targeting with `policytree`.
:::

------------------------------------------------------------------------

# Key Takeaways {.center}

## Summary

1. **CATEs** ($\tau(x)$) capture treatment effect heterogeneity — the ATE is just an average of CATEs

2. **Naive subgroup analysis** overfits — honest estimation is essential for valid HTE inference

3. **Causal trees** use honest splitting (separate samples for structure and estimation)

4. **Causal forests** aggregate many honest trees, with local centering (same DR scores from Unit 2) and forest-based inference

5. **`grf` package** provides causal forests with automatic honesty, local centering, and valid CIs

6. **Always test:** `test_calibration()` before interpreting heterogeneity patterns

------------------------------------------------------------------------

## What's Next

::: {.callout-note appearance="simple"}
## Coming Up

**Monday Mar 30:** Meta-Learners (T, S, X, R, DR) + **Jigsaw 3**

- Read your assigned Jigsaw 3 paper (posted on Canvas)
- Papers on causal forest applications in health economics

**Problem Set 3** assigned next week — HTE estimation and meta-learner comparison (due Apr 8)
:::

. . .

**Prep for Meta-Learners:**

- Chernozhukov et al. (2025) — Ch. 8: Heterogeneous Treatment Effects
- Nie & Wager (2021) — "Quasi-Oracle Estimation of Heterogeneous Treatment Effects" (*Biometrika*)

------------------------------------------------------------------------

## Readings for Today's Topics

**Required:**

- Wager (2024) — [Chapter 6: Causal Forests](http://web.stanford.edu/~swager/causal_inf_book.pdf)

**Recommended:**

- Athey & Imbens (2016) — "Recursive Partitioning for Heterogeneous Causal Effects" (*PNAS*)
- Athey, Tibshirani & Wager (2019) — "Generalized Random Forests" (*Annals of Statistics*)
- Wager & Athey (2018) — "Estimation and Inference of Heterogeneous Treatment Effects using Random Forests" (*JASA*)
- Davis & Heller (2017) — "Using Causal Forests to Predict Treatment Heterogeneity" (*AER P&P*)

------------------------------------------------------------------------

## Questions?

**Office hours:** By appointment — email or Canvas message
