---
title: Lab N
subtitle: Subtitle
author: Sean Sylvia
date: "2024-12-31"
slug: lab-0
categories: 
    -  Lab
    -  TOPIC
description: "Description"
format: html
    code-background: true
execute:
    warning: false
    echo: false
draft: true
---

# Lab Exercise

## Overview
This lab session focuses on **[Insert Topic Here]**. The goal of this exercise is to apply the concepts covered in class to practical problems using R.

---

## Learning Objectives
By the end of this lab, you will be able to:
1. [Insert Learning Objective Here]
2. [Insert Additional Learning Objective Here]
3. [Insert Additional Learning Objective Here]

---

## Case Scenario
Welcome to today’s lab! You will take on the role of **[Insert Role Here]**, tasked with solving a real-world problem related to **[Insert Scenario Description Here]**. This scenario will guide the exercises and provide a context for applying your skills.

---

## Instructions
1. Download the dataset from [Insert Link Here].
2. Follow the steps outlined in the exercises below.
3. Submit your completed lab exercise in Gradescope [Link] by [Insert Deadline Here].

---

## Exercises
### Exercise 1: [Insert Exercise Title Here]
- **Objective**: [Insert Objective Here]
```{r}
# Step 1: [Insert Description Here]
# Example Code
[Insert R Code Here]
```

### Exercise 2: [Insert Exercise Title Here]
- **Objective**: [Insert Objective Here]
```{r}
# Step 1: [Insert Description Here]
# Example Code
[Insert R Code Here]
```

---

## Reflection Questions
Answer the following questions as part of your lab report:
1. [Insert Question Here]
2. [Insert Additional Question Here]
3. [Insert Additional Question Here]

---

## Submission Guidelines
- Submit your R script and a brief write-up of your results.
- Include answers to the reflection questions.

**Deadline**: [Insert Date and Time]

---

## Notes
Ensure that your code is well-commented and your results are clearly presented. Use ggplot2 or other visualization libraries for graphical outputs where applicable.
